### 2011-02-10 doc/MVSeg.tex
ENH: updating atropos and MVSeg doc

@@

\section{Methods} Our goal will be to provide reproducible
measurements derived from multivariate segmentation.
\begin{description}
\item[RBG:] The principal direction of the tensor can be visualized in
  RGB space.  Some segmentation methods rely on this feature 
  modulated by the FA.
\item[FA:] Fractional anisotropy derived from the tensor.  This is
  equal to the variance of the eigenvalues of the tensor matrix.  It
  is the most common measure used in DTI studies, along with
  tractography. 
\item[Tractography:] White matter tracts provide a physical connection
  between and within cortical regions.  At the coarse scale of DTI, we mostly
  look at connections between cortical regions.  A simple way to
  define a tract from DTI is to follow the principal eigenvector until
  some stopping criterion based on curvature or FA is reached. 
\end{description}
An EM segmentation method can use any of these features.  One project
that may be similar to our own is here
\url{http://www.nitrc.org/projects/dots} though there is no clear
reference other than the website.  Prior DT segmentation work is here
\citep{Lenglet2005} and here \citep{Awate2007}.  Classic work on
partial volume estimation \citep{Santago1993,Shattuck2001}.

@@

>>>>

\subsection{EM image segmentation, briefly} 
Consider a classic image segmentation problem where one wants to
automatically label each image voxel into one of three classes, $A, B, C$.  We
use a Bayesian model for the probability of each part of the image being a member of a
specific class, $Q$, 
$$
P( Q | i ) \propto P ( i | Q ) P( Q ) 
$$
where $i$ is the intensity.
We then model 
$$ P( i | Q )  \propto \exp( - \| i - \mu_Q \|^2 / \sigma_Q^2 ) $$ 
where $\mu_Q , \sigma_Q$ are the parameters to a Gaussian describing
the intensity distribution of class $Q$.  For now, let us assume that
$P(Q)=\pi_q$ where $\pi_q$ is some constant between 0 and 1.  EM will use this
type of model for each class and, given initial values for the
parameters (the means and variances for each class's Gaussian), find
the maximum a posteriori solution to the class labels and parameters.
It's basically a gradient descent algorithm.  In this case, one simple EM
solution will be 
\begin{enumerate}
\item Take initial values for the class parameters from the user.
\item For each voxel in the image, compute the posterior probability
  for each label, that is, $P(A|i), P(B|i), P(C|i)$.  Assign the voxel
  to the label with maximal probability.  This is a synchronous
  updating scheme.  
\item For each class, recompute the parameters to the Gaussian by 
$ \mu_Q \leftarrow \frac{\sum_{j=1}^N i P(Q|i)}{\sum_{j=1}^N P(Q|i)}$
and $ \sigma_Q \leftarrow \frac{\sum_{j=1}^N (i-\mu_Q)^2 P(Q|i)}{\sum_{j=1}^N P(Q|i)}$.
\item If converged, stop, otherwise go to 2.  
\end{enumerate}
This is the basic idea of EM.  Seems simple.  However, it gets quickly complicated when
you begin to specify EM for a problem domain.  For instance, for
convergence guarantees, one would use iterated conditional modes (ICM)
in step 2.  A scientist should use his or her domain knowledge to
define the prior probabilities $P(Q)$ and to define the distance
(here, $\| i -\mu_Q \|$) and/or likelihood model in
$P(i|Q)$.  A common goal is to enforce local sets of voxels to have the same label, if possible.  Markov
random field frameworks provide many options for achieving label regularity 
in image segmentation \citep{Besag1986}.  One might also take care in estimating the mixing proportions that are
used in a standard finite mixture model.  This is all discussed in more detail
in the Atropos paper.  In the current work, we are concerned mainly
with the likelihood model and the input data representations which should be
chosen such that they are suitable for deriving biological and/or anatomical meaning from DTI.

\subsection{Diffusion tensor primer} A diffusion tensor (DT) is a
symmetric positive definite matrix.  Diffusion tensor imaging derives
a dense field of DT estimates at each position in a three-dimensional
imaging volume by measuring the amount of magnetically-induced water
diffusion in each of $N$ directions.  Directions, here, refers to
positions on the sphere (e.g. longitude and latitude but only in the
northern hemisphere).  Each tensor may be described uniquely as
$$D=\sum_i \lambda_i e_i^T e_i $$ where $e_i$ is a unit 3-vector and
$\lambda_i$ is a scalar in 0 to 1.  The $\lambda_i$ defines the
relative amount of diffusion along the axis defined by $e_i$.  A
highly anisotropic tensor will have $\lambda_1 >> \lambda_{2,3} $, that
is, the prinicipal eigenvector of the tensor will have much higher
diffusion than the other eigenvectors.   The variance of the
$\lambda_i$ is known as the fractional anisotropy (FA) and their mean is
known as the mean diffusion, apparent diffusion coefficient (ADC) or
the trace (actually the sum). Below are some of other measurements derived from DTI:
\begin{description}
\item[RBG:] The principal direction of the tensor can be visualized in
  RGB space.  Some segmentation methods rely on this feature 
  modulated by the FA.  That is, the RBG image voxel may be defined as
  $\vec{rgb}=[R=255~\text{FA}~e_1^1, G=255~\text{FA}~e_1^2, B=255~\text{FA}~e_1^3$].
\item[FA:] Fractional anisotropy derived from the tensor.  This is
  equal to the variance of the eigenvalues of the tensor matrix.  It
  is the most common measure used in DTI studies, along with
  tractography. 
\item[Tensor shape:]  This is determined by the tensor eigenvalues.  As
  neuroscientists interested in tracts, we may be concerned with how
  collections of eigenvalues are distributed along tracts and how
  these distributions differentiate tracts or tissues.
\item[Tensor orientation:] The principal
  eigenvector can point to any position on the upper hemisphere.  As
  communicated by the power of the RGB display of tensor information,
  this is a powerful feature. 
\item[Tractography:] White matter tracts provide a physical connection
  between and within cortical regions.  At the coarse scale of DTI, we mostly
  look at connections between cortical regions.  A simple way to
  define a tract from DTI is to trace a curve that follows the principal eigenvector until
  some stopping criterion based on curvature or FA is reached. 
\end{description}
So, in this work, we'd like to use these different features to help us
automatically differentiate tissues and tracts in not only population
average images but also individual images.  Being able to compute
these types of measures will let us do specific, novel studies of
neural architecture across species, gender, age and disease.

\subsection{A quick tensor segmentation example}
atropos seg example with dti

two questions --- one of representation , one of distance.  that is,
distances on what?
\begin{figure}
\begin{center}
\begin{tabular}{c}
\includegraphics[width=6in]{Figures/macaque_rgb.pdf}
\end{tabular}
\caption{We perform an RGB-based segmentation of the macaque
  population average tensor image.  At left is the manually placed
  labels, as performed in ITK-Snap.  At right is the Atropos result.
  You may repeat this study by labeling the macaque image with labels
  1, 2 and 3 (R, G, B) similarly to this example.  You then
  save the segmentation and run the script entitled MVSeg/src/scripts/seg\_dti\_manual\_init\_atropos.}
\label{fig:example1}
\end{center}
\end{figure}

@@

\section{Methods} Our goal will be to provide reproducible
measurements derived from multivariate segmentation.  An EM
segmentation method can use any of the DTI features we discussed
above.  One project that may be similar to our own is here
\url{http://www.nitrc.org/projects/dots} though there is no clear
reference other than the website.  Prior DT segmentation work is here
\citep{Lenglet2005} and here \citep{Awate2007}.  Classic work on
partial volume estimation \citep{Santago1993,Shattuck2001}.

\subsection{Ideas}
One idea to explore is nonparametric joint histogram based likelihood
models for tensor shape and orientation.  $J_{orientation} = f(i,j)/
\sum_{ij} f(i,j) $ where $f(i,j)$ is the frequency histogram of the
upper hemisphere describing orientation.  two parameters: theta and
psi.  $J_{shape} = g(i,j)/ \sum_{ij} g(i,j) $ where $g(i,j)$ is the
frequency histogram describing the co-occurrence of first and second
eigenvalues.  These would become our likelihood models (recalling
$P(i|Q)$) and would be initialized by priors deformed from previously
segmented images.

@@

\subsection{Code}
The class called {\em antsHistogramParzenWindowsListSampleFunction}
implements a non-parametric histogram based likelihood function.  It is located in MVSeg/src/ANTS/ImageSegmentation.
One might implement a class called  {\em
  antsJointHistogramParzenWindowsListSampleFunction} that could
implement the idea above.   Or one could follow the example {\em
  antsLogEuclideanGaussianListSampleFunction} to implement a
parametric likelihood.  Each class should be tested as a stand-alone
method and should be verified to work before being included in the
segmentation functionality.
